---
title: "Safaricom Digital Academy Assignments Week 3:-Data Visualisation with ggplot2"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# AfterWork Data Science: Data Visualisation with Python Project:- Emmanuel Odenyire Anyira

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the research question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?

### Installing and loading libraries
```{r}
install.packages("tidyverse")
library(tidyverse)
install.packages('readr')
library(readr)
install.packages("dplyr")
library(dplyr)
install.packages("knitr")
install.packages("rmarkdown")
library(dplyr)
library(rmarkdown)
install.packages("reshape")
library(reshape)
```


## 2. Reading the Data

```{r}



# Load the data below
# --- 
# This is the dataset used for this assignment
# Dataset url = https://bit.ly/2WvQbhJ
# We use read.csv to read and load data in R
# --- 
# YOUR CODE GOES BELOW
# 
hotel_bookings_df <- read.csv("hotel_bookings.csv", header = TRUE, na.strings=c("NULL",NA))

```

```{r}
# Checking the first 5 rows of data
# We use head() function to check the first 5 rows
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_bookings_df, 5)
```

```{r}
# Checking the last 5 rows of data
# We use the tail() function o check the last 5 records
# ---
# YOUR CODE GOES BELOW
# 
tail(hotel_bookings_df, 5)
```

```{r}
# Sample 10 rows of data
#We use the sample() to check 10 samples
# ---
# YOUR CODE GOES BELOW
# 
sample_n(hotel_bookings_df, 10)
```

```{r}
# Checking number of rows and columns
# We use the dim() function to check the number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotel_bookings_df)
```

```{r}
# Checking datatypes
# We use the glimpse() function to check the data types in the data frame
# ---
# YOUR CODE GOES BELOW
# 
glimpse(hotel_bookings_df)
```

```{r}

```

Record your general observations below:

Observation 1:- We have 119390 rows/records and 32 columns
Observation 2: - 'Integer' is the common data type


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# We use str() to check data types in R and is.na() to check NULL entries
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_bookings_df)
sum(is.na(hotel_bookings_df))
```

We observe the following from our dataset:

Observation 1:- Int and Chr are the two common data types
Observation 2:- There are 478 missing values in our data set

```{r}
# Checking how many duplicate rows are there in the data
# We use duplicated() to check for duplicates
# ---
# YOUR CODE GOES BELOW
# 
# hotel_booking_df[duplicated(hotel_booking_df), ]
sum(duplicated(hotel_bookings_df))
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 
which(colSums(is.na(hotel_bookings_df))  == ncol(hotel_bookings_df))
```

We observe the following from our dataset:

Observation 1:- There are no columns in which all the data is null
Observation 2:- We have 31994 duplicates

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW

which(rowSums(is.na(hotel_bookings_df)) == nrow(hotel_bookings_df))
```

We observe the following from our dataset:

Observation 1:  there is no row in which all the data is null
Observation 2

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_bookings_df)

# Select the desired columns
hotel_bookings_df2 <- hotel_bookings_df %>%
  na.omit() %>%
  select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, adults, children, babies, is_repeated_guest, previous_cancellations, previous_bookings_not_canceled, days_in_waiting_list, total_of_special_requests)
hotel_bookings_df2

# Create a correlation matrix
corr_matrix <- cor(hotel_bookings_df2, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black") +
  geom_text(
    aes(label = round(value, 2)),
    color = "white"
  ) +
  coord_fixed() +
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    limit = c(-1,1)
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
hotel_bookings_df = select(hotel_bookings_df, -company)
head(hotel_booking_df)
```

From the data variable description we see that the Distribution Channel category that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

hotel_bookings_df$agent[is.na(hotel_bookings_df$agent)] <- "TO"
hotel_bookings_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

hotel_bookings_df <- subset(hotel_bookings_df, adults!=0 | children!=0 | babies!=0)
hotel_bookings_df
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
hotel_bookings_df$children[is.na(hotel_bookings_df$children)] <- round(mean(hotel_bookings_df$children, na.rm = TRUE), 0)
hotel_bookings_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
sum(is.na(hotel_bookings_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
# hotel_booking_df$children <- lapply(hotel_booking_df$children, as.numeric)
as.integer(hotel_bookings_df$children)
hotel_bookings_df
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

hotel_bookings_df %>%
  group_by(reservation_status) %>%
  summarise(is_canceled = sum(is_canceled))

ggplot(hotel_bookings_df, aes(x= reservation_status, y= is_canceled))+
  geom_col(stat = "identity", width =  0.5, fill= "blue") +
  labs(
    title = "Reservation Status",
    x= "Status",
    y= "Number"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Ariel Black", color="purple")
  )
  
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

resort_city_hotel <- hotel_bookings_df %>%
  count(hotel)
resort_city_hotel

ggplot(data=resort_city_hotel, aes(x="", y=n, fill=hotel)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") +
  geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette="Set2")+
  labs(
    fill = "Hotel",
    title = "Booking Between Resort Hotel and City Hotel"
  )+
  theme_void()
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# 

year_booking <- hotel_bookings_df %>%
  count(arrival_date_year)
year_booking

ggplot(data=year_booking, aes(x="", y=n, fill=factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "blue"
  ) +
  coord_polar("y") +
    geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +
  labs(
    fill = "Years",
    title = "Booking for each year") +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
busy_months <- hotel_bookings_df %>%
  count(arrival_date_month)
busy_months

ggplot(busy_months, aes(x= arrival_date_month, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "magenta") +
  labs(
    title = "Bookings per Month",
    x= "Month",
    y= "Number of bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#

hotel_bookings_df

countries <- hotel_bookings_df %>%
  group_by(country) %>%
  summarise(guests = sum(adults)+sum(children)+sum(babies)) %>%
  # arrange(desc(guests)) %>%
  top_n(3)
countries

ggplot(countries, aes(x= country, y= guests))+
  geom_col(stat = "identity", width =  0.8, fill= "blue") +
  labs(
    title = "Top 3 Countries per Guests",
    x= "Countries",
    y= "Number of Guests"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="black")
  )+
  coord_flip()
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 

stay_length <- hotel_bookings_df %>%
  group_by(hotel) %>%
  summarise(total_stay_days = sum(stays_in_weekend_nights)+sum(stays_in_week_nights))
stay_length

ggplot(data = stay_length, aes(x= hotel, y= total_stay_days))+
  geom_col(stat = "identity", width =  0.8, fill= "grey") +
  labs(
    title = "Days Stayed per Hotel",
    x= "Hotel Type",
    y= "Number of Days Stayed"
  ) +
  theme(
    plot.title = element_text(hjust = 0, size=16, family="Arial", color="blue")
  )+
  theme_classic()
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation <- hotel_bookings_df %>%
  count(reserved_room_type)
accommodation

ggplot(data = accommodation, aes(x= reserved_room_type, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "brown") +
  labs(
    title = "Bookings per Accomodation Type",
    x= "Room Type",
    y= "Number of Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )+
coord_flip()
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 

```

Our observations:
-

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
b). Do we need other data to answer our question?
c). Did we have the right question?


